Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC and DLPFC with TMM normlaisation and SVA)

### 2017-05-19 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_SVA.Rmd
Major modifications to account for new model selection with SVA

cl = makeCluster(14)
registerDoParallel(cl)

@@

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (ACC)
COUNT_ACC_ID = 'syn7217917';
ALL_USED_IDs = COUNT_ACC_ID
COUNT_ACC = read.table(synGet(COUNT_ACC_ID)@filePath, header=T, sep='\t', check.names = F, nrows = 56632) %>%
  dplyr::select(-chr, -start, -end, -strand, -length)

# Download reprocessed counts (DLPFC)
COUNT_DLPFC_ID = 'syn7217916';
ALL_USED_IDs[length(ALL_USED_IDs)+1] = COUNT_DLPFC_ID
COUNT_DLPFC = read.table(synGet(COUNT_DLPFC_ID)@filePath, header=T, sep='\t', check.names = F, nrows = 56632) %>%
  dplyr::select(-chr, -start, -end, -strand, -length)

COUNT = join_all(list(COUNT_ACC, COUNT_DLPFC), type = 'full')
rownames(COUNT) = COUNT$geneID
COUNT$geneID = NULL
COUNT = data.matrix(COUNT)

# Get merged metadata
METADATA_ID = 'syn2299154'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = F)

@@

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
                          "Dx", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
                          'DNA_genotyping_report: Genotyping Sample_ID')

# Filter metadata 
ind = as.character(METADATA[1,] %>% unlist) %in% metadataVarOfInterest
METADATA = METADATA[,ind]
colnames(METADATA) = gsub('[^[:alnum:]]','_',as.character(METADATA[1,] %>% unlist))

METADATA_ACC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, ACC_RNA_isolation__Exclude_,
                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_,
                ACC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_, 
                LibraryBatch = ACC_RNA_report__Library_Batch,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'ACC')

METADATA_DLPFC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, DLPFC_RNA_isolation__Exclude_,
                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_,
                DLPFC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                LibraryBatch = DLPFC_RNA_report__Library_Batch,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC')

METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY)

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Individual_ID", "Institution", "Gender", "LibraryBatch", "Dx.Tissue")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- METADATA$SampleID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```

Correlation/association between covariates at an FDR <= 0.1

```{r covariates.correlation, fig.width=8, fig.height=8}
ind.cov = setdiff(colnames(COVARIATES), 'Individual_ID')
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES[,ind.cov], PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
```

Explore metatdata

```{r dataExplore, fig.width = 6, fig.height = 15}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

@@

multiplot(plotlist = p, cols = 1)
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx.Tissue), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% unique()

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to NA
log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix[log.mat] = NA

# TMM normalisation
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(VOOM.GENE_EXPRESSION$E))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Dx.Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="top")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Institution', 'Dx.Tissue')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Institution', 'Dx.Tissue')]))
```

```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                 COVARIATES[,ind.cov],
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

>>>>

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

@@

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```

### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (ACC)
COUNT_ACC_ID = 'syn8414424'
ALL_USED_IDs = COUNT_ACC_ID
COUNT_ACC = downloadFile(COUNT_ACC_ID)

# Download reprocessed counts (DLPFC)
COUNT_DLPFC_ID = 'syn8413221';
ALL_USED_IDs[length(ALL_USED_IDs)+1] = COUNT_DLPFC_ID
COUNT_DLPFC = downloadFile(COUNT_DLPFC_ID)

# Join ACC and DLPFC together
COUNT = join_all(list(COUNT_ACC, COUNT_DLPFC), type = 'full') %>%
  as.data.frame()

GENE.LEN = dplyr::select(COUNT_ACC, Geneid, Length) %>%
  dplyr::rename(ensembl_gene_id = Geneid) %>%
  unique()
rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id

rownames(COUNT) = COUNT$Geneid
ind = setdiff(colnames(COUNT), c('Geneid', 'Chr', 'Start', 'End', 'Strand', 'Length'))
COUNT = data.matrix(COUNT[,ind])

@@

# Get RNA qc metrics from synapse
METADATA_QC_ACC_ID = 'syn2929053'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_ACC_ID
METADATA_QC_ACC = read.table(synGet(METADATA_QC_ACC_ID)@filePath, fill = T, header = T, sep = ',') %>%
  lapply(as.character)

# Get RNA qc metrics from synapse
METADATA_QC_DLPFC_ID = 'syn2279446'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
METADATA_QC_DLPFC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',') %>%
  lapply(as.character)

METADATA_QC = rbindlist(list(METADATA_QC_ACC, METADATA_QC_DLPFC), use.names = T, fill = T) %>%
  dplyr::filter(is.na(Exclude.)) %>%
  dplyr::select(Sample.RNA.ID, Ribozero.Batch, Library.Batch, Flowcell.Batch, Mapped.Reads, 
                Intragenic.Rate, Exonic.Rate, Intronic.Rate, Intergenic.Rate, Genes.Detected, 
                Transcripts.Detected, Expression.Profiling.Efficiency, rRNA.Rate, Total.Reads, Percent.Aligned) %>%
  dplyr::rename(SampleID = Sample.RNA.ID, RibozeroBatch = Ribozero.Batch, LibraryBatch = Library.Batch, 
                FlowcellBatch = Flowcell.Batch, MappedReads = Mapped.Reads, IntragenicRate = Intragenic.Rate, 
                ExonicRate = Exonic.Rate, IntronicRate = Intronic.Rate, IntergenicRate = Intergenic.Rate, 
                GenesDetected = Genes.Detected, TranscriptsDetected = Transcripts.Detected, 
                ExpProfEfficiency = Expression.Profiling.Efficiency, rRNARate = rRNA.Rate, 
                TotalReads = Total.Reads, PercentAligned = Percent.Aligned)

# Get merged metadata
METADATA_ID = 'syn4484273'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = T, check.names = F)

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", "Dx",

                          "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?",

                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?",

                          "DNA_genotyping_report: Genotyping Sample_ID")

# Filter metadata 
ind = colnames(METADATA) %in% metadataVarOfInterest
METADATA = METADATA[,ind]
colnames(METADATA) = gsub('[^[:alnum:]]','_',colnames(METADATA))

METADATA_ACC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                DNA_genotyping_report__Genotyping_Sample_ID, ACC_RNA_isolation__Exclude_, 
                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'ACC')

METADATA_DLPFC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                DNA_genotyping_report__Genotyping_Sample_ID, DLPFC_RNA_isolation__Exclude_, 
                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC')

METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
  droplevels() %>%
  list(.,METADATA_QC) %>%
  join_all %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY) %>%
  unique

@@

```{r gene.param}
## Get GC content from biomart
backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content", "gene_biotype"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::select(ensembl_gene_id, percentage_gc_content) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$ensembl_gene_id
```

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){
  x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- METADATA$SampleID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], function(x){
  x = as.numeric(as.character(gsub('[\\,\\%]','',x)))
})

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
tmp = COVARIATES.CORRELATION$ESTIMATE
tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
draw(h, heatmap_legend_side = 'left')
```

### Explore metatdata
```{r data.explore, fig.width = 12, fig.height = 12}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))

@@

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme

# Transcripts Detected
p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TranscriptsDetected)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Transcripts Detected') + my.theme

# Mapped Reads
p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme

# PercentAligned
p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PercentAligned)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('Percent Aligned') + my.theme

# rRNARate
p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot()
p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme

multiplot(plotlist = p, cols = 3)

# Institution
# vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES)

# Gender
# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx.Tissue), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC.CONT$ensembl_gene_id[!is.na(GENE.GC.CONT$percentage_gc_content)]) %>%
  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$Length)])

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
  dplyr::rename(ensembl_gene_id = genes) %>%
  left_join(Ensemble2HGNC) %>%
  group_by(gene_biotype) %>%
  summarise(fraction  = n()) %>%
  filter(fraction > 100) %>%
  dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
```
`r dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]` genes are used for the analysis.

Following is the distribution of biotypes
`r kable(pct.pc)`

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
# Based on PC
indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")

# Based on number of gene outliers
indToRemove = c(indToRemove, "MSSM_RNA_ACC_BP_15", "MSSM_RNA_ACC_175", "MSSM_RNA_ACC_224", "MSSM_RNA_ACC_240", 
                "MSSM_RNA_ACC_285", "MSSM_RNA_ACC_344", "MSSM_RNA_ACC_376", "MSSM_RNA_ACC_374", "PENN_RNA_ACC_003", 
                "PITT_RNA_ACC_539", "MSSM_RNA_PFC_201", "MSSM_RNA_PFC_234", "MSSM_RNA_PFC_355", "MSSM_RNA_PFC_389",
                "MSSM_RNA_PFC_80", "MSSM_RNA_PFC_83", "PENN_RNA_PFC_22", "PENN_RNA_PFC_59", "PENN_RNA_PFC_62", 
                "PENN_RNA_PFC_80", "PENN_RNA_PFC_105", "PITT_RNA_PFC_1472", "PITT_RNA_PFC_566", "PITT_RNA_PFC_727", 
                "PITT_RNA_PFC_802")

# # Outliers from Menahcem's DLPFC analysis syn3493755
# outliers = c('MSSM_RNA_BP_PFC_1', 'MSSM_RNA_PFC_21', 'MSSM_RNA_BP_PFC_3', 'MSSM_RNA_PFC_103', 
#              'MSSM_RNA_PFC_105', 'MSSM_RNA_PFC_115', 'MSSM_RNA_PFC_169',  'MSSM_RNA_PFC_17',
#              'MSSM_RNA_PFC_18', 'MSSM_RNA_PFC_19', 'MSSM_RNA_PFC_213', 'MSSM_RNA_PFC_214', 
#              'MSSM_RNA_PFC_215', 'MSSM_RNA_PFC_216', 'MSSM_RNA_PFC_217', 'MSSM_RNA_PFC_218',
#              'MSSM_RNA_PFC_219', 'MSSM_RNA_PFC_222', 'MSSM_RNA_PFC_24', 'MSSM_RNA_PFC_252', 
#              'MSSM_RNA_PFC_259', 'MSSM_RNA_PFC_263', 'MSSM_RNA_PFC_267', 'MSSM_RNA_PFC_275',
#              'MSSM_RNA_PFC_307', 'MSSM_RNA_PFC_322', 'MSSM_RNA_PFC_325', 'MSSM_RNA_PFC_27', 
#              'MSSM_RNA_PFC_332', 'MSSM_RNA_PFC_333', 'MSSM_RNA_PFC_334', 'MSSM_RNA_PFC_335',
#              'MSSM_RNA_PFC_340', 'MSSM_RNA_PFC_35', 'MSSM_RNA_PFC_362', 'MSSM_RNA_PFC_372', 
#              'MSSM_RNA_PFC_373', 'MSSM_RNA_PFC_385', 'MSSM_RNA_PFC_393', 'MSSM_RNA_PFC_395',
#              'MSSM_RNA_PFC_49', 'MSSM_RNA_PFC_54', 'MSSM_RNA_PFC_70', 'MSSM_RNA_PFC_76', 
#              'MSSM_RNA_PFC_77', 'MSSM_RNA_PFC_78', 'MSSM_RNA_PFC_88', 'MSSM_RNA_PFC_93',
#              'FAKE_DLPFC_CMC_MSSM_426', 'PENN_RNA_PFC_29', 'FAKE_DLPFC_CMC_PENN_055', 
#              'PENN_RNA_PFC_75', 'PENN_RNA_PFC_110', 'FAKE_DLPFC_CMC_PENN_106', 'FAKE_DLPFC_CMC_PENN_107',
#              'FAKE_DLPFC_CMC_PENN_108', 'MSSM_RNA_PFC_14', 'MSSM_RNA_BP_PFC_21', 'MSSM_RNA_PFC_293', 
#              'MSSM_RNA_PFC_349', 'MSSM_RNA_PFC_353', 'MSSM_RNA_PFC_363', 'MSSM_RNA_PFC_369',
#              'PENN_RNA_PFC_82', 'PENN_RNA_PFC_89', 'PITT_RNA_PFC_1159', 'PITT_RNA_PFC_1296', 'PITT_RNA_PFC_537', 
#              'PITT_RNA_PFC_567', 'PITT_RNA_PFC_871', 'PITT_RNA_PFC_878', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139',
#              'MSSM_RNA_PFC_210',  'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_278', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_348',
#              'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_56', 'PITT_RNA_BP_PFC_10003', 'PITT_RNA_BP_PFC_1047',
#              'PITT_RNA_BP_PFC_1086', 'PITT_RNA_BP_PFC_1092', 'PITT_RNA_BP_PFC_1247', 'PITT_RNA_BP_PFC_1282', 
#              'PITT_RNA_BP_PFC_1324', 'PITT_RNA_BP_PFC_1374', 'PITT_RNA_BP_PFC_1391', 'PITT_RNA_BP_PFC_686')

# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
  dplyr::mutate(label = SampleID)
plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

# Plot abberent distribution of logcpm counts
tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% indToRemove),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
p

indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
COVARIATES = COVARIATES[indToRetain,]

tmp = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue'), sep = '_') %>%
  group_by(Dx, Tissue) %>%
  summarise(count = n()) %>%
  spread(Tissue, count)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals

Based on the expression pattern following `r length(indToRemove)` samples were tagged as outliers based on their mrna expression: `r paste(indToRemove, collapse = ', ')`

Distribution of samples are:
`r kable(tmp)`

### Identify gene outliers
Set outlier expression of genes in each samples to NA
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(LOG.CPM, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = NA

# Distribution of fraction NA genes in samples
tmp.frac = apply(log.mat, 2, sum)/dim(LOG.CPM)[1]
writeLines('Distribution of fraction NAs in samples')
quantile(tmp.frac, seq(0, 1, 0.1)) %>%
  round(3) 
top.nsamples = names(tmp.frac)[tmp.frac >= 0.02]
```
Only `r length(top.nsamples)` samples have more than 2% of NA genes. They are: `r paste(top.nsamples, collapse = ', ')`
